---
title: "class11"
format: gfm
---

# Section 1. Proportion of G/G in a population 

Downloaded a CSV file from Ensemble

Here we read this CSV file
```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```
```{r}
table(mxl$Genotype..forward.strand.)
```
```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl)*100
```

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```

Find proportion of G|G
```{r}
round(table(gbr$Genotype..forward.strand.)/nrow(gbr)*100, 2)
```

This variant is associated with childhood asthma is frequent in the GBR population than the MXL population. 
Let's now dig into this further

## Section 4: Population scale analysis 

How many samples do we have?
```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```
```{r}
nrow(expr)
```
```{r}
table(expr$geno)
```
```{r}
library(ggplot2)
```
Let's make a boxplot
```{r}
ggplot(expr)+
  aes(x=geno, y=exp, fill=geno)+
  geom_boxplot(notch = TRUE)
```
Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

A|A is expressed more than G|G. The A|A genotype is correlated with more expression of ORMDL3.

