---
title: "Class 17: Mini Project"
author: "Christina Mac"
format: pdf
---
We will start by downloading the most recently dated "Statewide COVID-19 Vaccines Administered by ZIP code" CSV file:
```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

Q1. What column details the total number of people fully vaccinated? 
Column 11, "persons_fully_vaccinated"

Q2. What column details the Zip code tabulation area?
zip_code_tabulation_area

Q3. What is the earliest date in this dataset? 
2021-01-05

```{r}
head(sort(vax$as_of_date))
```

Q4. What is the latest date in this dataset?
2023-02-28

```{r}
head(sort(vax$as_of_date, decreasing = TRUE))
```

We can use the skim() function for a quick overview of a new dataset like this: 

```{r}
skimr::skim(vax)
```

Q5. How many numeric columns are in this dataset? 
13

Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?
16525

```{r}
sum(is.na(vax$persons_fully_vaccinated))
```

What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 
8.3%

```{r}
length(vax$persons_fully_vaccinated)
round((sum(is.na(vax$persons_fully_vaccinated))/length(vax$persons_fully_vaccinated))*100, 2)
```
##Working with dates 

the lubridate package makes working with dates and times in R much less of a pain. Let's have a first play with this package here.

```{r}
library(lubridate)
```
```{r}
today()
```
```{r}
# This will give an Error!
#today() - vax$as_of_date[1]
```


```{r}
# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```
We can now magically do math with dates

```{r}
today()-ymd("2021-01-05")
```

How old am I?
```{r}
today()-ymd("2001-02-11")
```

Let's treat the whole col

How many days have passed since the first vaccination reported in this dataset?

```{r}
today()-ymd(vax$as_of_date[1])
```

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

Q9. How many days have passed since the last update of the dataset?
7 days

```{r}
today()-ymd(vax$as_of_date[nrow(vax)])
```

Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 
113 dates

```{r}
length(unique(vax$as_of_date))
```

# Working with ZIP codes

ZIP codes are also rather annoying things to work with as they are nmeric but no in the conventional sense of doing math. 

Just like dates we have speial packages to help us work with ZIP codes.

```{r}
library(zipcodeR)
```
```{r}
geocode_zip('92037')
```
```{r}
zip_distance('92037','95148')
```
```{r}
reverse_zipcode(c('92037', "95148") )
```

# Focus on the San Diego area
Let's now focus on the San Diego county 

```{r}
# Subset to San Diego county only areas
sd <- vax[ vax$county == "San Diego" , ]
```

```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```
```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```


```{r}
head(sd.10)
```


Q11. How many distinct zip codes are listed for San Diego County? 
107 zipcodes 

```{r}
length(unique(sd$zip_code_tabulation_area))
```

Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset? 
92154

```{r}
sd$zip_code_tabulation_area[which.max(sd$age12_plus_population)]
```

```{r}
x <-sd %>%
  filter(as_of_date == "2023-02-28")
```


Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-02-28”? 
74%

```{r}
sd.today <- filter(sd, as_of_date == "2023-02-28")
```
```{r}
mean(sd.today$percent_of_population_fully_vaccinated, na.rm = T)
```


Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-02-28”?


```{r}
library(ggplot2)
```
```{r}
ggplot(sd.today, aes(x=sd.today$percent_of_population_fully_vaccinated, )) + 
  geom_histogram(bins = 10, na.rm=T) +
  labs(x= "Percent of Population Fully Vaccinated in a Zip COde Area", y="Count (Zip code areas") +
  ggtitle("Histogram of Vaccination Rates Across San Diego County")
```



```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```


Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ggplot(ucsd) +
  aes(x=as_of_date,
      y=percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(title = "Vaccination rate for La Jolla CA 92037", x= "Date", y="Percent Vaccinated")
```


```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2022-11-15")

head(vax.36)
```

Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-11-15”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
average_percent <- mean(vax.36$percent_of_population_fully_vaccinated)
```


```{r}
ggplot(ucsd) +
  aes(x=as_of_date,
      y=percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(title = "Vaccination rate for La Jolla CA 92037", x= "Date", y="Percent Vaccinated") +
  geom_hline(yintercept = average_percent,color = "red", linetype=10)
```

Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-11-15”?


```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```
Q18. Using ggplot generate a histogram of this data.
```{r}
ggplot(vax.36) +
  aes(x=percent_of_population_fully_vaccinated) +
  geom_histogram() +
  labs(x= "Percent Vaccinated", y="Count") +
  xlim(0,1)
```

Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?
Both averages are lower 

```{r}
x <- filter(vax.36, zip_code_tabulation_area %in% c("92109", "92040"))
x$percent_of_population_fully_vaccinated
```


```{r}
vax %>% filter(as_of_date == "2022-11-15") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```
```{r}
vax.36.all <- filter(vax,age5_plus_population > 36144)


ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="purple", na.rm = T) +
  ylim(0,1) +
  labs(x="Date", y="Percent Vaccinated",
       title= " Vaccination rate across California",
       subtitle="Only areas with a population above 36k are shown.") +
  geom_hline(yintercept = average_percent, linetype=5)
```
Q21. How do you feel about traveling for Thanksgiving Break and meeting for in-person class afterwards?
Would be tiring, but I would be okay with it!
