---
title: "class10"
format: pdf
---
```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```
Q1. How many different candy types are in this dataset?
85 diferent candy types

```{r}
nrow(candy)
```

Q2. How many fruity candy types are in the dataset?
38 fruit candy types
```{r}
sum(candy$fruity)
```
```{r}
candy["Twix", ]$winpercent
```
Q3. What is your favorite candy in the dataset and what is it’s winpercent value?
M&M's winpercent is 66.57458%
```{r}
candy["M&M's", ]$winpercent
```

Q4. What is the winpercent value for “Kit Kat”?
76.7686%
```{r}
candy["Kit Kat", ]$winpercent
```

Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?
49.6535%
```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
```{r}
library("skimr")
skim(candy)
```
Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
It seems like n_missing and complete_rate look like they are on a binary scale instead of a continuous scale like the other columns are.

Q7. What do you think a zero and one represent for the candy$chocolate column?
This means that it either has or does not have chocolate in it.


Q8. Plot a histogram of winpercent values
```{r}
hist(candy$winpercent)
```

Q9. Is the distribution of winpercent values symmetrical?
No, the distribution is right skewed.

Q10. Is the center of the distribution above or below 50%?
Below

Q11. On average is chocolate candy higher or lower ranked than fruit candy?
Higher
```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])

```

Q12. Is this difference statistically significant?
Yes, it is statistically significant since the p-value is less than 0.05.
```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)],candy$winpercent[as.logical(candy$fruity)])
```
Q13. What are the five least liked candy types in this set?
Nik L Nip, Boston Baked Beans, Chiclets, Super bubble, Jawbusters
```{r}
head(candy[order(candy$winpercent),], n=5)
```

Q14. What are the top 5 all time favorite candy types out of this set?
Reese's Peanut Butter cup, Reese's Miniatures,Twix, Kit Kat, Snickers
```{r}
tail(candy[order(candy$winpercent),], n=5)
```
Q15. Make a first barplot of candy ranking based on winpercent values.
```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_bar(stat="identity")
```
Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?
```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_bar(stat="identity")
```
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```
```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```
Q17. What is the worst ranked chocolate candy?
Sixlets

Q18. What is the best ranked fruity candy?
Starburst

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = Inf)
```
Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?
Fruity

Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?
Nik L Kip (least popular), Nestle Smarties, Ring pop, Hershey's Krackel, Hershey's Milk Chocolate
```{r}
tail(candy[order(candy$pricepercent),], n=7)
```
```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
```{r}
library(corrplot)
```
```{r}
cij <- cor(candy)
corrplot(cij)
```
Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
Chocolaty and Fruity are the most anti-correlated.

Q23. Similarly, what two variables are most positively correlated?
Chocolaty and Bar are the most positively correlated.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```
```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = Inf)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
```{r}
library(plotly)
```

```{r}
#ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?
Yes since they all were inexpensive and had moderate to low winpercent.Fruity, hard, and pluribus
